---
title: "Fixed Income"
author: "Kyle McWreath"
date: "8/13/2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This application illustrates the effects of default risk on the returns from holding bonds to maturity. The expected return on a bond is different from the bond's promised return. I use a Markov model to solve for the expected rate of return on a risky bond. The process takes into account the probability of default, the transition of the issuer from one state of credit worthiness to another and the percentage recovery of face value when the bond defaults.

## Calculating The Expected Return in a One-Period Framework

* $F$  = face value of a bond
* $P$  = price of a bond
* $Q$  = annual coupon rate of the bond
* $\pi$  = probability that the bond will not default at end of year
* $\lambda$ = fraction of bond's value the bondholder will collect upon default

 The bond's expected end of year cash flow is calculated as: 
$\pi*(1+Q)*F+(1-\pi)*\lambda*F$

The one year expected return of the bond is calculated as= $\frac {\pi*(1+Q)*F+(1-\pi)*\lambda*F}{P}-1$

Below you can analyze the relationship between the variables of the bond and the expected cash flow:




```{r echo=FALSE}
inputPanel(
  numericInput("fv", label = "Choose the face value:",
              value = 100, min = 1, max = 100, width = '50%', step= 1),
    numericInput("price", label = "Choose the price of the bond:",
              value = 95, min = 1, max = 500, width = '50%', step = 1),
  
  numericInput("q", label = "Choose the annual coupon rate of the bond:",
              value = .5, min = 0, max = 1, width = '50%', step = .05),
  
   numericInput("pi", label = "Choose the probability that the bond will not default:",
              value = .5, min = 0, max = 1, step = .01, width = '50%'),
  numericInput("lam", label = "Fraction of bond's value bondholders collect upon default:",
              value = .5, min = 0, max = 1, step = .01, width = '50%')

)

```
```{r echo = FALSE} 
DT:: renderDataTable({
  
  x <- input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv
  y = ((input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv)/input$price)-1
  
  
  x <- matrix(c(x,y),nrow = 1, ncol = 2)
 
  DT:: datatable(x, colnames =c("Expected Cash Flow at End of Year 1", "Expectd Return %"), escape = FALSE, options = list(dom = 't'))
 
             
                       
})
  ```


##Calculating a Bond's Expected Return in a Multi-Period Framework
Now we use a basic Markov model that utilizes a ratings transition matrix to compute a bonds expected return
  
Assuming a simple ratings system: 
  
 
$$ \Pi =
\begin{vmatrix}
\pi_{AA}&\pi_{AB}&\pi_{AC}&\pi_{AD}&0\\
\pi_{BA}&\pi_{BB}&\pi_{BC}&\pi_{BD}&0\\
\pi_{CA}&\pi_{CB}&\pi_{CC}&\pi_{CD}&0\\
0&0&0&0&1\\
0&0&0&0&1\\
\end{vmatrix}
$$
    
The probabilities in each row of the matrix $\Pi$ indicate the probability that in one period the bond will go from a rating of $i$ to $j$

In order to calculate the expected return on a risky bond we must assume default probabilities of each of the ratings, A,B,C,D and E.    
    
    
```{r, results='asis', echo=FALSE}
  
  new <- matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","   E")),2), c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0,0,0,.22,.0,0,0,0,0,1,1))
  
knitr::kable(new, full_width = F)
```

The table above indicates that if a bond is rated A in the current period, there is a probability of 0.97 that it will still be rate A in the next period.  

There is a probability of 0.02 that it will be rated B in the next period and a probability of 0.01 that it will be rated C.  

It is impossible for a bond to be rated A today and D or E in the subsequent period.   

A bond that starts off with a rating of B can in a subsequent period be rated A with a probability of 0.05, be rated B with a probability of 0.8 or be rated C with a probability of 0.15  

The transition probabilities from state C to states A, B, C are 0.01, 0.02, 0.75, 0.22.  

While it is possible to go from ratings A, B, C to any of ratings A, B, C, or D, it is not possible to go from A, B, C to E. This is true because E denotes that default took place in the previous period.  

A bond that is currently in state D (first time default) will necessarily be in E in the next period. This is why the fourth row of matrix 
the matrix described above will always be:
$$
 The~5th~row~of~matrix~ \Pi = \begin{vmatrix}
0&0&0&0&1\\
 \end{vmatrix}
$$  
Once the rating is in E it remains there permanently. 

##The Multi-Period Transtion Matrix  
The matrix described above defines the transition probabilities in one period. 
The two period transition probabilities are given by the matrix product $\Pi *\Pi$.  
The interactive table below shows the two year transition probability of default and can be adjusted for any number of periods: 

```{r}

##define the trasnstion probabilities over 1 period.
period1 <-matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","E")),2), 
c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0, 0,0,.22,.0,0, 0,0,0,1,1))

period1
##The use of the defined Mmatrix function below provides a simple way to calculate the default probabilites of a given bond in any number of periods
Mmatrix <- function(matrix, n) {
  

 if (n == 1){
    print(matrix)
 } 
  else { matrix%^%n
  }
}

##Testing the function above for a 2 period transition probability matrix:
library(expm)
Mmatrix(period1,2)
```  
The two period transition matrix above implies that if a bond is rated "B" today, there is a probability of 9.4% that in two periods it will be rated "A", a probability of 64.4% that in two periods it will be rated B, a probability of 23.3% that in two periods it will be rated C and a probability of 3.3% that in two periods it will default (rated D)  

In general, the year $t$ transition matrix is given by the matrix power $\Pi{^t}$:  


```{r echo=FALSE}
shinyUI(fluidPage(

sidebarLayout(
  
  sidebarPanel(
   tags$style("[type = 'number'] {font-size:18px;}"),
   numericInput("year", label = "Choose the number of periods:",
              value = 3, min = 1, max = 100, width = '1000%', step= 1)
    
   ),

mainPanel(
  dataTableOutput("year")
 
   ) 
  )
 )
)
```
```{r echo = FALSE} 
  
  period1 <-matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","E")),2), 
c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0, 0,0,.22,.0,0, 0,0,0,1,1))
  
Mmatrix <- function(matrix, n) {
  

 if (n == 1){
    print(matrix)
 } 
  else { matrix%^%n
  }
}

DT:: renderDataTable({
  
    DT:: datatable(Mmatrix(period1, input$year),escape = FALSE, options = list(dom = 't'), caption = paste(c(input$year, "- Period Transition Probability Matrix"),collapse = " "))
 
             
                       
})
  ```  
  
If you set the above transition matrix to 10 periods, you can see that if a bond started out with an A rating, there is a probability of 1.48% that the bond will be in default at the end of 10 periods and a 7.06% probability that it will have defaulted in a previous period (rated E).

  
## The Bond Payoff Vector  
  
Above we defined $Q$ as the bonds coupn rate and $\lambda$ as the recovery percentage payoff of face value if the bond defaults.  
The payoff vector of the bond depends on whether the bond is currently in its last period $N$ or whether $t < N$.  

    
\[ Payoff(t,t<N) = \left( \begin{array}{cc}
Q\\
Q\\
Q\\
\lambda\\
0\\
\end{array} 
%
\right)
Payoff(t, T = N) = \left(\begin{array}{cc}
1 + Q \\
1 + Q \\
1 + Q \\
\lambda \\
0 \\
\end{array} \right)
\]  
* The first three elements of each vector denote the payoff in non-defaulted states  
* The fourth element \lambda is the payoff if the rating is D  
* The fifth element  0 is the payoff if the bond rating is E(E is the rating for the period after the bond defaults)  

The initial state payoff vector

* The initial state vector is a vector with a 1 for the current rating of the bond and zeros elsewhere
* If the bond has a rating of A at date 0 then: 
  $$
 initial = \begin{vmatrix}
1&0&0&0&0\\
 \end{vmatrix}
$$   
* If the bond has a rating of B at date 0 then:
  $$
 initial = \begin{vmatrix}
0&1&0&0&0\\
 \end{vmatrix}
$$  
  
    
##Expected Bond Payoff in Period $t$:  
  
  $E[Payoff(t)] = Initial* \Pi{^t}*Payoff(t)$  
  
  #Initial state vector
  initialstateofbond <- function(state1) {
      
      if (state1 == 'A') {
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(1,0,0,0,0))
      }
      else if (state1=='B'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,1,0,0,0))
      }
      else if (state1=='C'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,1,0,0))
      }
      
      else if (state1=='D'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,0,1,0))
      }
      else if (state1=='E'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,0,0,1))
      }
      
      
    } 
###TO Do Create App that evaluates all inputs and plots outputs
