---
title: "Fixed Income"
author: "Kyle McWreath"
date: "8/13/2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This applicaiton illustrates the effects of default risk on the returns from holding bonds to matruity. The expected return on a bond is differnt from the bond's promised retun. I use a Markov model to solve for the expected rate of return a risky bond. The process takes into account the probability of default, the transition of the issuer from one state of credit worthiness to another and the percentage recovery of face value when the bond defaults.

## Calculating The Expected Return in a One-Period Framework

* $F$  = face value of a bond
* $P$  = price of a bond
* $Q$  = annual coupon rate of the bond
* $\pi$  = probibility that the bond will not default at end of year
* $\lambda$ = fraction of bond's valuethe bondholder will collect upon default

 The bond's expected end of year cash flow is calculated as: 
$\pi*(1+Q)*F+(1-\pi)*\lambda*F$

The one year expected return of the bond is calculated as= $\frac {\pi*(1+Q)*F+(1-\pi)*\lambda*F}{P}-1$

Below you can analyze the realtionship between the varaibles of the bond and the expected cash flow:




```{r echo=FALSE}
inputPanel(
  numericInput("fv", label = "Choose the face value:",
              value = 100, min = 1, max = 100, width = '50%', step= 1),
    numericInput("price", label = "Choose the price of the bond:",
              value = 95, min = 1, max = 500, width = '50%', step = 1),
  
  numericInput("q", label = "Choose the annual coupon rate of the bond:",
              value = .5, min = 0, max = 1, width = '50%', step = .05),
  
   numericInput("pi", label = "Choose the probability that the bond will not default:",
              value = .5, min = 0, max = 1, step = .01, width = '50%'),
  numericInput("lam", label = "Fraction of bond's value bondholders collect upon default:",
              value = .5, min = 0, max = 1, step = .01, width = '50%')

)

```
```{r echo = FALSE} 
DT:: renderDataTable({
  
  x <- input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv
  y = ((input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv)/input$price)-1
  
  
  x <- matrix(c(x,y),nrow = 1, ncol = 2)
 
  DT:: datatable(x, colnames =c("Expected Cash FLow at End of Year 1", "Expectd Return %"), escape = FALSE, options = list(dom = 't'))
 
             
                       
})
  ```


##Calculating a Bond's Expected Return in a Multi-Period Framework
Now we use a basic Markov model that utilizes a ratings transition matrix to compute a bonds expected return
  
* Assuming a simple ratings system: 
  
 
$$ \Pi =
\begin{vmatrix}
\pi_{AA}&\pi_{AB}&\pi_{AC}&\pi_{AD}&0\\
\pi_{BA}&\pi_{BB}&\pi_{BC}&\pi_{BD}&0\\
\pi_{CA}&\pi_{CB}&\pi_{CC}&\pi_{CD}&0\\
0&0&0&0&1\\
0&0&0&0&1\\
\end{vmatrix}
$$
    
* The probabilities in each row of the matrix $\Pi$ indicate the probability that in one period the bond will go from a rating of $i$ to $j$

    
    
    
```{r, results='asis', echo=FALSE}
  
  new <- matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","   E")),2), c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0,0,0,.22,.0,0,0,0,0,1,1))
  
knitr::kable(new, full_width = F)
```


##TO DO: Add the tranisition matrix function that I built last weekend and create reactive output
