---
title: "Bond Analysis"
author: "Kyle McWreath"
date: "8/20/2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This summary provides some basic analytical tools to help understand the effects of default risk on returns from holding risky bonds to maturity. I provide a few simple applications that illustrate that the expected return on a bond is different from the bond's promised return. The following concepts are covered:   

* Calculating the expected return on a bond in a one-period framework  
* A Markov transition model to analyze the default risk of a bond with an initial rating  
* Creating a bond payoff vector based on the timing of payoff  
* The expected payoff of a bond in any given period of it's life

## Calculating the Expected Return in a One-Period Framework  

To calculate the expected return in a one-period framework, the following equations can be used:

 The bond's expected end of year cash flow is calculated as: 
$\pi*(1+Q)*F+(1-\pi)*\lambda*F$

The one year expected return of the bond is calculated as= $\frac {\pi*(1+Q)*F+(1-\pi)*\lambda*F}{P}-1$  
Where:

* $F$  = face value of a bond
* $P$  = price of a bond
* $Q$  = annual coupon rate of the bond
* $\pi$  = probability that the bond will not default at end of year
* $\lambda$ = fraction of bond's value the bondholder will collect upon default



Below you can analyze the relationship between the variables of the bond and the expected cash flow:




```{r echo=FALSE}
inputPanel(
  numericInput("fv", label = "Face value:",
              value = 100, min = 1, max = 100, width = '50%', step= 1),
    numericInput("price", label = "Price of the bond:",
              value = 100, min = 1, max = 500, width = '50%', step = .5),
  
  numericInput("q", label = "Annual coupon rate of the bond %:",
              value = .05, min = 0, max = 1, width = '50%', step = .01),
  
   numericInput("pi", label = "Probability that the bond will not default%:",
              value = .9, min = 0, max = 1, step = .01, width = '50%'),
  numericInput("lam", label = "% of bond's value bondholders collect upon default:",
              value = .05, min = 0, max = 1, step = .01, width = '50%')

)

```
```{r echo = FALSE} 
DT:: renderDataTable({
  
  x <- input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv
  y = ((input$pi*(1+input$q)*input$fv+(1-input$pi)*input$lam*input$fv)/input$price)-1
  
  
  x <- matrix(c(x,y*100),nrow = 1, ncol = 2)
 
  DT:: datatable(x, colnames =c("Expected Cash Flow at End of Year 1", "Expectd Return %"), escape = FALSE, options = list(dom = 't'))
 
             
                       
})
  ```


##Calculating a Bond's Expected Return in a Multi-Period Framework
Now we use a basic Markov model that utilizes a ratings transition matrix to compute a bonds expected return. More information on Markov models can be found here [Markov Modeling](http://www.mathpages.com/home/kmath232/part2/part2.htm).
  
Assuming a simple ratings system of A, B, C, D, E: 
  
 
$$ \Pi =
\begin{vmatrix}
\pi_{AA}&\pi_{AB}&\pi_{AC}&\pi_{AD}&0\\
\pi_{BA}&\pi_{BB}&\pi_{BC}&\pi_{BD}&0\\
\pi_{CA}&\pi_{CB}&\pi_{CC}&\pi_{CD}&0\\
0&0&0&0&1\\
0&0&0&0&1\\
\end{vmatrix}
$$
    
The probabilities in each row of the matrix $\Pi$ indicate the probability that in one period the bond will go from a rating of $i$ to $j$.

In order to calculate the expected return on a risky bond we must assume default probabilities of each of the ratings, A,B,C,D and E.    
    
    
```{r, results='asis', echo=FALSE}
  
  new <- matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","   E")),2), c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0,0,0,.22,.0,0,0,0,0,1,1))
  
knitr::kable(new, format = "markdown", align = 'l')
```

The table above indicates that if a bond is rated A in the current period, there is a probability of 0.97 that it will still be rated A in the next period.  

There is a probability of 0.02 that it will be rated B in the next period and a probability of 0.01 that it will be rated C.  

It is impossible for a bond to be rated A today and D or E in the subsequent period.   

A bond that starts off with a rating of B can in a subsequent period be rated A with a probability of 0.05, be rated B with a probability of 0.8 or be rated C with a probability of 0.15.      
Bonds rated B in the current period do not default (rating D) in the next period.

The transition probabilities from state C to states A, B, C are 0.01, 0.02, 0.75, 0.22.  

While it is possible to go from ratings A, B, C to any of ratings A, B, C, or D, it is not possible to go from A, B, C to E. This is true because E denotes that default took place in the previous period.  

A bond that is currently in state D (first time default) will necessarily be in E in the next period. This is why the fourth row of matrix 
the matrix described above will always be:
$$
 The~5th~row~of~matrix~ \Pi = \begin{vmatrix}
0&0&0&0&1\\
 \end{vmatrix}
$$  
Once the rating is in E it remains there permanently. 

##The Multi-Period Transtion Matrix  
The matrix described above defines the transition probabilities in one period. 
The two period transition probabilities are given by the matrix product $\Pi *\Pi$.  
The interactive table below shows the two year transition probability of default and can be adjusted for any number of period.  
The following R code demonstrates how to set up the Markov transition matrix.

```{r}

##define the trasnstion probabilities over 1 period.
period1 <-matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","E")),2), 
c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0, 0,0,.22,.0,0, 0,0,0,1,1))

period1
##The use of the defined Mmatrix function below provides a simple way to calculate the default probabilites of a given bond in any number of periods
Mmatrix <- function(matrix, n) {
  

 if (n == 1){
    print(matrix)
 } 
  else { matrix%^%n
  }
}

##Testing the function above for a 2 period transition probability matrix:
library(expm)
Mmatrix(period1,2)
```  
The two period transition matrix above implies that if a bond is rated "B" today, there is a probability of 9.4% that in two periods it will be rated "A", a probability of 64.4% that in two periods it will be rated B, a probability of 23.3% that in two periods it will be rated C and a probability of 3.3% that in two periods it will default (rated D)  

In general, the year $t$ transition matrix is given by the matrix power $\Pi{^t}$:  


```{r echo=FALSE}
shinyUI(fluidPage(

sidebarLayout(
  
  sidebarPanel(
   tags$style("[type = 'number'] {font-size:18px;}"),
   numericInput("year", label = "Choose the number of periods:",
              value = 10, min = 1, max = 100, width = '1000%', step= 1)
    
   ),

mainPanel(
  dataTableOutput("year")
 
   ) 
  )
 )
)
```
```{r echo = FALSE} 
  
  period1 <-matrix(nr= 5, nc = 5, dimnames = rep(list(rating = c("A","B","C","D","E")),2), 
c(.97,.05,.01,0,0,.02,.8,.02,0,0,.01,.15,.75,0,0, 0,0,.22,.0,0, 0,0,0,1,1))
  
Mmatrix <- function(matrix, n) {
  

 if (n == 1){
    print(matrix)
 } 
  else { matrix%^%n
  }
}

DT:: renderDataTable({
  
    DT:: datatable(Mmatrix(period1, input$year),escape = FALSE, options = list(dom = 't'), caption = paste(c(input$year, "- Period Transition Probability Matrix"),collapse = " "))
 
             
                       
})
  ```  
  
If you set the above transition matrix to 10 periods, you can see that if a bond started out with an A rating, there is a probability of 1.48% that the bond will be in default at the end of 10 periods and a 7.06% probability that it will have defaulted in a previous period (rated E).

  
## The Bond Payoff Vector  
  
Above we defined $Q$ as the bonds coupn rate and $\lambda$ as the recovery percentage payoff of face value if the bond defaults.  
The payoff vector of the bond depends on whether the bond is currently in its last period $N$ or whether $t < N$.  

    
\[ Payoff(t,t<N) = \left( \begin{array}{cc}
Q\\
Q\\
Q\\
\lambda\\
0\\
\end{array} 
%
\right)
Payoff(t, T = N) = \left(\begin{array}{cc}
1 + Q \\
1 + Q \\
1 + Q \\
\lambda \\
0 \\
\end{array} \right)
\]  

* The first three elements of each vector denote the payoff in non-defaulted states  
* The fourth element \lambda is the payoff if the rating is D  
* The fifth element  0 is the payoff if the bond rating is E(E is the rating for the period after the bond defaults)  



 
## Initial State of the Bond Vector  
  In order to cacluate expected payoffs we need to define the initial state of the bond vector. This vector is a vector with a 1 for the  current rating of the bond an zeros elsewhere. An example  is provided below:
  
  
* The initial state vector is a vector with a 1 for the current rating of the bond and zeros elsewhere
* If the bond has a rating of A at date 0 then: 
  $$
 initial = \begin{vmatrix}
1&0&0&0&0\\
 \end{vmatrix}
$$   
* If the bond has a rating of B at date 0 then:
  $$
 initial = \begin{vmatrix}
0&1&0&0&0\\
 \end{vmatrix}
$$  
  
  ```{r echo=FALSE}
inputPanel(
  selectInput("rating", "Choose initial rating:",
                   choices = c("A", "B", "C","D","E"), width = '50%')
       

)

```
```{r echo = FALSE} 

output$rating = DT:: renderDataTable({
  

             
                       
})
  ```
  
  ```{r echo = FALSE}
  
   initialstateofbond <- function(state1) {
      
      if (state1 == 'A') {
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(1,0,0,0,0))
      }
      else if (state1=='B'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,1,0,0,0))
      }
      else if (state1=='C'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,1,0,0))
      }
      
      else if (state1=='D'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,0,1,0))
      }
      else if (state1=='E'){
        nep <- matrix(nr= 5, nc = 1, dimnames = rep(list(rating = c("A","B","C","D","E")),1), c(0,0,0,0,1))
      }
      
      
    } 
DT:: renderDataTable({
 
 x<- initialstateofbond(input$rating)

DT:: datatable(x,escape = FALSE, options = list(dom='t'), colnames = c("State"))             
                       
})
  ```
    
    
##Expected Bond Payoff in Period $t$:  
  
The followng equation provides the soultion for the expected payoff at any discrete time in the bond's life 
  
  $E[Payoff(t)] = Initial* \Pi{^t}*Payoff(t)$  

Where:  
* $Initial$ = the  initial  state  vector    
* $\Pi{^t}$ = the transition matrix in period $t$  
* $Payoff(t)$ = the payoff vector in time $t$


  
  
  


  
  
  
